---
title: "Biden Projected to Win in the 2024 US Presidential Election"
author:
  - Monica Sainani
  - Kaavya Kalani
thanks: "Code and data supporting this analysis is available at: https://github.com/sainanim/Predictions-for-the-US-2024-Elections"
date: today
date-format: long
abstract: "This study investigates the dynamics of voting behaviour of americans leveraging demographic, socioeconomic, and geographic factors, such as age, family income, education, state of residency, and sex. We use a Bayesian Logistic Regression model with Post-Stratification to predict the outcome between Biden and Trump with the mentioned factors as predictors. Our study indictates that Biden is likely to be re-elected as the president with a 10% lead. This research contributes to a better understanding of the factors influencing political engagement in American democracy by offering insightful information that might guide future electoral strategies and policy discussion."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(rstan)
library(tidyverse)
library(arrow)
library(dplyr)
library(knitr)
library(ggplot2)
library(gutenbergr)
library(haven)
library(labelled)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(broom.mixed)
library(here)
library(kableExtra)

# read the datasets and model
poststrat_data <- arrow::read_parquet(file = here("data/analysis_data/poststrat_data.parquet"))
survey_data <- arrow::read_parquet(file = here("data/analysis_data/survey_data.parquet"))
us_election_model <- readRDS(file = here("models/single_bay.rds"))
```

# Introduction
A key turning point in the political environment of the United States occurred with the 2020 presidential election, which was marked by extreme conflict and serious economic and social challenges. We analyze data from the American Community Survey (ACS) 2022 sample, which offers a thorough understanding of demographic, socioeconomic, and geographic aspects across the nation, in order to comprehend the patterns underlying voting choices during this 2024 election period.

Our estimand is the relationship between demographic, socio-economic, and geographic variables (such as age, income, education, state of residence, and gender) and voting behavior during the 2022 presidential election. Using this extensive dataset, our goal is to identify trends and factors that influence voting behaviour, clarifying the complex relationship between personal traits and local factors that influences election results.

Our research adds to a greater understanding of the factors influencing voting decisions through comprehensive
statistical analysis, yielding insightful data that may influence future election campaigns and policy discussions. Through highlighting the subtleties of voting patterns in the 2022 election, our research provides a critical insight on the condition of American democracy and the factors motivating involvement in politics and engagement.

The paper is further organised into three sections: Data, Model, Results, and Discussion. In the Data section, we discuss how the datasets to be used for the analysis were obtained and pre-processed. We will explain the variables of interest in the datasets for the analysis. The Model section describes the model being used for the analyis. The Results section will then highlight and discuss the trends and associations found during the analysis. Lastly, the Discussion section will talk about some interesting trends found in Results in depth, link it to the real world and also highlight the weaknesses of our analysis.

# Data

##Introduction to the Data:
We will use the most recent American Community Survey, the 2022 Sample, as the post-stratification dataset
to represent the voter population and adjust the weight of the survey data. The population for this sample was the entire US population, categorized into households as subgroups, and the responses of the sampled subgroups were collected. ACS 2022 dataset consists of 3,373,378 observations.

Using a 1-in-100 national random sample, the ACS 2022 sample provides a thorough overview of the demographic,
socioeconomic, and geographic features of the US population. Most importantly, this dataset includes people who live in both home and group quarters, offering a comprehensive view of the American population.

Since the ACS 2022 sample is weighted, the data are guaranteed to accurately reflect the demographic makeup of the country's population. Weighting improves the reliability and accuracy of statistical analyses and conclusions generated from the dataset by facilitating more precise extrapolation of findings from the sample to the larger population.

The ACS 2022 sample is noteworthy for its geographic structure, where the Public Use Microdata Area (PUMA) is the
smallest recognizable geographic unit.
Geographic areas that do not cross state boundaries and have a population of at least 100,000 are known as PUMAs.
This physical separation makes it easier to analyze regional differences and trends in socioeconomic and demographic traits, enabling scholars to investigate the subtleties of voting patterns and electoral dynamics throughout the nation.

All things considered, the ACS 2022 sample is a useful tool for statistical analysis and empirical study, providing insights into the complex aspects of American culture. In the sections that follow, we explore the variables that make up the dataset and consider how important they are to comprehending voting patterns and electoral results for the 2024 election using this 20022 survey data.

## Introduction to the Variables
These are the key variables utilized in our statistical analysis of voting behavior during the 2022 survey.
These variables are drawn from the American Community Survey (ACS) 2022 sample and encompass demographic,
socio-economic, and geographic characteristics that may influence individuals' voting preferences.

Age: Age is a representation of each candidate's actual age as of the election. A key demographic factor that might affect voting patterns, generational dynamics, and political opinions is age.

Income: The pre-tax amount of the respondents' families' entire annual income from all sources is shown as income. As a stand-in for socioeconomic position and financial security, income has the power to influence people's political choices, attitudes, and worldviews.

Education: Based on the highest year of school or degree earned, respondents' educational attainment is indicated by their education. knowledge electoral dynamics requires a knowledge of the important variable of educational attainment, which is directly connected to political ideology, civic participation, and voter turnout.

State: The respondents' state of residence provides information about where their housing units are located and makes it possible to analyze local politics and voting trends. The political climate, the priorities of policies, and the demographic makeup of a state can all affect how people vote and how elections turn out.

Sex: The report of sex indicates a person's gender. Political beliefs and actions are greatly influenced by gender dynamics, which may have an impact on voting patterns and election results.

## Model
We used a Bayesian Logistic Regression model with Post-Stratification to predict the 2024 US Presidential Elections. Logistic regression is a method used for binary classification to predict the probability of a categorical dependent variable. 

For our analysis, a logistic regression model will be first used to model the proportion of the voters who will vote for Biden. The model will be based on five independent demographic variables: state, sex, age, education and income.

The logistic regression model we will be using is:
\begin{equation}
\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{state} + \beta_2 \times \text{sex} + \beta_3 \times \text{age} + \beta_4 \times \text{education} + \beta_5 \times \text{income} \label{eq:model1}
\end{equation}

where,

- $\hat{p}$ represents the probability that someone will vote for Biden
- $\beta_0$ represents the intercept term of this logistical regression. It is the probability that someone will vote for Biden if the predictors' values are zero
- $\beta_1$ is the coefficient corresponding to state
- $\beta_2$ is the coefficient corresponding to sex
- $\beta_3$ is the coefficients corresponding to age
- $\beta_4$ is the coefficients corresponding to education
- $\beta_5$ is the coefficients corresponding to income

In our model, normal priors with a mean of 0 and a standard deviation of 2.5 are used for both the coeffients and the intercept. Setting the mean of the priors to 0 implies that there is no expectation of a particular direction or magnitude for the coefficients or intercept. We chose this as we have no expectation of the same. The standard deviation of 2.5 reflects the uncertainty or variability in the prior beliefs. We chose a moderately wide prior to allow for a reasonable amount of uncertainty. 

The chosen priors allow the data to largely determine the posterior distribution as they are relatively non-informative. They don't heavily influence the results unless the data provide strong evidence to the contrary.

The use of moderately wide priors can also help regularize the model, preventing overfitting and providing more stable estimates, particularly when dealing with limited data.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-model-coefficients
#| tbl-cap: Coefficients of the model

# Create model summary
model_summary <- modelsummary(us_election_model)

# Print model summary in a nice format
print(model_summary)

# Extract coefficients
coefficients <- broom::tidy(us_election_model, conf.int = TRUE)
coefficients |> kable(digits = 2)
```

@tbl-model-coefficients shows the coefficients for our model that can be used in the equation along with the standard error and the 95% confidence interval. The standard error (SE) is a measure of the precision with which a sample statistic estimates a population parameter. It quantifies the variability of sample statistics around the population parameter. A 95% confidence interval means that if you were to take many samples and construct confidence intervals in the same way, approximately 95% of those intervals would contain the true population mean.

After we created the model, we then post-stratify these proportions and to conclude who we predict to win the 2024 US Presidential Elections. Post-stratification is a statistical technique employed to refine survey estimates by aligning them with known population characteristics. It helps us ensure that our findings accurately represent the diversity of the entire population. 

In this process, we first the organize the survey data based on the demographic factors we decided to investigate. This involved counting individuals in each group to understand the sample distribution and computing proportions to depict the relative contribution of each demographic group to the overall survey sample.

Predictive model is then employed to generate probabilities for each demographic group, capturing the nuanced political preferences within distinct strata. These probabilities were summarized by aggregating them within each demographic group. 

Applying this post-stratification process results in enhanced predictions that align with the known population distribution within each stratum. These adjusted values contribute to a more accurate representation of party
votes within specific demographic categories, ensuring our findings are reflective of the broader population.


## Results
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-support-state
#| fig-cap: Estimated proportion support for Biden by State

# Calculate proportions by state
proportions_state <- poststrat_data %>%
  group_by(state, sex, age, education, income) %>%
  summarise(n = n()) %>%
  group_by(state) %>%
  mutate(prop = n/sum(n))

# Obtain predictions from the Bayesian model
proportions_state$estimate <- predict(us_election_model, newdata = proportions_state, type = "response")
proportions_state <- proportions_state %>% mutate(num_voters = n*estimate)

# Calculate prediction errors and intervals
errors <- predict(us_election_model, newdata = proportions_state, type = "response", se.fit = TRUE)
upper_state <- errors$fit + errors$se.fit
lower_state <- errors$fit - errors$se.fit
proportion_error_state <- cbind(proportions_state, lower_bound = lower_state, upper_bound = upper_state) %>%
  rename(lower_bound = "lower_bound", upper_bound = "upper_bound")

proportion_error_state <- proportion_error_state %>%
  mutate(lower_estimate = n*lower_bound, upper_estimate = n*upper_bound)

# Calculate votes for Biden
vote_for_biden <- proportion_error_state %>%
  mutate(biden_predict_prop = estimate * prop,
         biden_predict_prop_lower = lower_bound * prop,
         biden_predict_prop_upper = upper_bound * prop) %>%
  group_by(state) %>%
  summarise(biden_predict = sum(biden_predict_prop),
            biden_lower = sum(biden_predict_prop_lower),
            biden_upper = sum(biden_predict_prop_upper))

# Prepare data for plotting
state_final_split <- survey_data %>%
  group_by(state, vote_biden) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

state_vote_for_biden <- inner_join(state_final_split, vote_for_biden, by = "state") %>%
  filter(vote_biden == '1')

# Create the plot
ggplot(state_vote_for_biden, aes(y = biden_predict, x = reorder(state, -biden_predict), color = "MRP estimate")) +
  geom_point() +
  geom_errorbar(aes(ymin = biden_lower, ymax = biden_upper), width = 0) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    x = "State",
    y = "Estimated proportion support for Biden",
    color = "Source"
  ) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  coord_flip() +
  theme(legend.position = "bottom")

```
@fig-support-state

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-support-age
#| fig-cap: Estimated proportion support for Biden by Age Groups

# Calculate proportions by age
proportions_age <- poststrat_data %>%
  group_by(state, sex, age, education, income) %>%
  summarise(n = n()) %>% 
  group_by(age) %>%
  mutate(prop = n/sum(n))

# Obtain predictions from the Bayesian model
proportions_age$estimate <- predict(us_election_model, newdata = proportions_age, type = "response")
proportions_age <- proportions_age %>% mutate(num_voters = n*estimate)

# Calculate prediction errors and intervals
errors <- predict(us_election_model, newdata = proportions_age, type = "response", se.fit = TRUE)
upper_age <- errors$fit + errors$se.fit
lower_age <- errors$fit - errors$se.fit
proportion_error_age <- cbind(proportions_age, lower_bound = lower_age, upper_bound = upper_age) %>% 
  rename(lower_bound = "lower_bound", upper_bound = "upper_bound")

proportion_error_age <- proportion_error_age %>%
  mutate(lower_estimate = n*lower_bound, upper_estimate = n*upper_bound)

# Calculate votes for Biden
vote_for_biden <- proportion_error_age %>%
  mutate(biden_predict_prop = estimate * prop,
         biden_predict_prop_lower = lower_bound * prop,
         biden_predict_prop_upper = upper_bound * prop) %>%
  group_by(age) %>%
  summarise(biden_predict = sum(biden_predict_prop),
            biden_lower = sum(biden_predict_prop_lower),
            biden_upper = sum(biden_predict_prop_upper))

# Prepare data for plotting
age_final_split <- survey_data %>%
  group_by(age, vote_biden) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

age_vote_for_biden <- inner_join(age_final_split, vote_for_biden, by = "age") %>%
  filter(vote_biden == '1')

# Plotting
ggplot(age_vote_for_biden, aes(x = age, y = biden_predict)) +
  geom_col(fill = "purple") +
  labs(title = "Proportions of Votes for Biden by age group",
       y = "Proportion of Votes",
       x = "Age groups")

```
@fig-support-age


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-support-sex
#| fig-cap: Estimated proportion support for Biden by Sex

# sex
# Calculate proportions by sex
proportions_sex <- poststrat_data %>%
  group_by(state, sex, age, education, income) %>%
  summarise(n = n()) %>% 
  group_by(sex) %>%
  mutate(prop = n/sum(n))

# Obtain predictions from the Bayesian model
proportions_sex$estimate <- predict(us_election_model, newdata = proportions_sex, type = "response")
proportions_sex <- proportions_sex %>% mutate(num_voters = n*estimate)

# Calculate prediction errors and intervals
errors <- predict(us_election_model, newdata = proportions_sex, type = "response", se.fit = TRUE)
upper_sex <- errors$fit + errors$se.fit
lower_sex <- errors$fit - errors$se.fit
proportion_error_sex <- cbind(proportions_sex, lower_bound = lower_sex, upper_bound = upper_sex) %>% 
  rename(lower_bound = "lower_bound", upper_bound = "upper_bound")

proportion_error_sex <- proportion_error_sex %>%
  mutate(lower_estimate = n*lower_bound, upper_estimate = n*upper_bound)

# Calculate votes for Biden
vote_for_biden <- proportion_error_sex %>%
  mutate(biden_predict_prop = estimate* prop,
         biden_predict_prop_lower = lower_bound * prop,
         biden_predict_prop_upper = upper_bound * prop) %>%
  group_by(sex) %>%
  summarise(biden_predict = sum(biden_predict_prop),
            biden_lower = sum(biden_predict_prop_lower),
            biden_upper = sum(biden_predict_prop_upper))

# Prepare data for plotting
sex_final_split <- survey_data %>%
  group_by(sex, vote_biden) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

sex_vote_for_biden <- inner_join(sex_final_split, vote_for_biden, by = "sex") %>%
  filter(vote_biden == '1')

# Plotting
ggplot(sex_vote_for_biden, aes(x = sex, y = biden_predict)) +
  geom_col(fill = "red") +
  labs(title = "Proportions of Votes for Biden by sex",
       y = "Proportion of Votes",
       x = "Sex groups")

```
@fig-support-sex


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-support-general
#| fig-cap: Estimated proportion support for Biden

general_proportion_data <- poststrat_data |>
  group_by(state, sex, age, education, income) |>
  summarise(n = n()) |>
  mutate(prop = n / sum(n))

# Predictions
general_proportion_data$estimate <- as.vector(predict(us_election_model, newdata = general_proportion_data))

# Prediction Intervals
general_errors <- as.data.frame(predict(us_election_model, newdata = general_proportion_data, se.fit = TRUE))
general_proportion_data$lower_bound <- general_errors$fit - general_errors$se.fit
general_proportion_data$upper_bound <- general_errors$fit + general_errors$se.fit

# Summarize predictions for Biden
general_vote_for_biden <- general_proportion_data %>%
  summarise(biden_predict = sum(estimate * prop),
            biden_lower = sum(lower_bound * prop),
            biden_upper = sum(upper_bound * prop))

# Plotting
general_data <- survey_data %>%
  group_by(vote_biden) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))

ggplot(general_data, aes(x = as.factor(vote_biden), y = prop)) +
  geom_col(fill = "blue") +
  labs(title = "Proportions of Votes",
       y = "Proportion",
       x = "Vote") +
  scale_x_discrete(labels = c("Vote for Biden", "Will not vote for Biden"))

```

@fig-support-general


# Discussion

## Insights Gained:

Our analysis, based on sample data from the 2022 survey, followed by post-stratification focusing on key variables, Age, State, and Sex, yielded valuable insights into voter behavior. Notably, our findings indicated an inclination within the population towards voting for Biden or the Democratic Party.

In the graph showcasing different regions and their voting for biden, we noticed that
This is similar to the findings in https://www.brookings.edu/articles/forecasting-the-presidential-election-what-can-we-learn-from-the-models/
that address the same prediction for 2024. Our model and the article predict that states that support Trump are
from the middle and Midwest parts of the US, whereas coastal states and the Northeast tend to vote for Biden.
Historically, rural, conservative-leaning towns with strong cultural ties to traditional values and industries like manufacturing and agriculture have defined the middle and Midwest regions of the United States.
These areas frequently support gun rights, minimal government intrusion, and religious liberty—issues that are strongly aligned with the agenda of the Republican Party, which Trump advocated. In contrast, the Northeast and coastline states, which are renowned for their progressive views on culture, urban facilities, and diverse populations, have historically tended toward more liberal policies and social ideals that are more in line with the Democratic Party platform, which Biden advocated.


We also created a graph highlighting different age groups and what their voting opinions are. We noticed that
the youngest group of voters (age 18-24). The emphasis on social justice and LGBTQ+ rights among the younger generation may be the cause of this difference. Since there are undoubtedly more young people who identify as sexual minorities, there is a trend for them to support Biden and the Democratic Party. We also noticed that the older generation i.e individuals of age group higher than 60 years old were inclined to voting for Biden as well.
National security and conservative principles, which are frequently connected to the Republican Party, may have an impact on the elder generation.

Lastly, our third variable of study was Sex. We noticed a slight bias towards the democratic party from the females than the male gender. Women are predicted to vote for the democratic party 0.6 times more than men.
Studies have consistently demonstrated that there is a gender difference in political ideology, with women generally favoring liberal or progressive policies over those of males. https://www.pewresearch.org/politics/2023/07/12/voting-patterns-in-the-2022-elections/#:~:text=In%20the%202022%20midterms%2C%2054,2022%20while%2051%25%20favored%20Democrats.
This discrepancy is sometimes linked to the fact that women prioritize issues more than men do, with women giving social welfare, healthcare, education, and reproductive rights—all of which are more in line with the Democratic Party program. With its focus on women's rights, healthcare reform, and inclusivity, Biden's campaign may have connected more deeply with female voters, resulting in higher levels of support.

## Comparative Analysis:
Our findings align with existing literature on electoral behavior, which underscores the significance of demographic variables in shaping voter preferences. While our study offers insights into the inclination towards voting for Biden, it is important to contextualize our findings within the broader landscape of political science research. https://projects.economist.com/us-2020-forecast/president/how-this-works


Our analysis contributes to a growing body of evidence highlighting the complex interplay between socio-economic factors and political attitudes, providing valuable insights for future research and decision-making in the political arena.

## Contributions to Knowledge: (Need to edit this)
By focusing our analysis on specific demographic variables, we contributed to a deeper understanding of the underlying determinants shaping electoral outcomes. Our study corroborates existing literature suggesting that demographic factors such as age, income, education, and geographic location play pivotal roles in influencing voter preferences. 
(https://projects.economist.com/us-2020-forecast/president/how-this-works)
Moreover, our utilization of post-stratification techniques allowed for a more granular examination of voting behavior, offering insights into the heterogeneity of voter sentiment across different demographic groups.


## Weaknesses and Limitations: (Need to edit this)

While our analysis provides valuable insights into voter behavior, it is important to acknowledge several limitations. 
Firstly, our reliance on sample data from the 2022 survey may introduce biases and inaccuracies in our predictions.
Our reliance on sample data from the 2022 survey may introduce a sampling bias, as the sample might not fully represent the diversity of the electorate. Consequently, this could lead to inaccuracies in our predictions and limit the generalizability of our findings to the broader population.

Additionally, while we focused on key demographic variables, other influential factors such as political ideology and campaign messaging were not fully accounted for in our model. This means that our research has the inaccuracy of Incomplete Variable Consideration.

Furthermore, uncertainties in the post-stratification process may impact the reliability of our estimates,
warranting caution in interpreting the observed inclination towards voting for Biden.

## Future Directions:
Moving forward, further research is warranted to address the limitations identified in our study and advance our
understanding of electoral dynamics. Future studies could explore the integration of additional data sources, such as social media sentiment analysis and polling data, to enhance the accuracy of election predictions.
Additionally, efforts to improve transparency and reproducibility in data collection and analysis are essential for enhancing the reliability and validity of electoral forecasts.

## Data Source:(Add where the second dataset is from)
The data utilized in this study were sourced from the 2022 survey,
supplemented by post-stratification data obtained from the American Community Surveys (ACS).
These datasets provided comprehensive information on voter demographics, socio-economic characteristics,
and geographic location, enabling a robust analysis of voting behavior in the 2020 US presidential election.

# Appendix

## Cleaning

## Survey dataset
Here is a glimpse of the survey data set used

```{r}
head(survey_data)
```

## Poststratification dataset
Here is a glimpse of the poststratification data set used
```{r}
head(poststrat_data)
```

# References